---
title: "Class 14 DESeq2"
author: "Peyton Chiu (PID:A18145937)"
format: html
editor: visual
toc: true
---

## Background

Here we are working through a RNASeq analysis with the input data coming from HOX genes

## Importing Data 

```{r}
colData <- read.csv("GSE37704_metadata.csv", row.names = 1)
countData<- read.csv("GSE37704_featurecounts.csv", row.names = 1)
```

```{r}
head(countData)
```

```{r}
head(colData)
```

We need to drop the column length

```{r}
countData <- as.matrix(countData[,-1])
head(countData)
```

We also need to filter genes with a 0 read count across all samples

```{r}
countData = countData[rowSums(countData) > 0, ]
head(countData)
```

The DeSeq object is set up

```{r}
library(DESeq2)

dds = DESeqDataSetFromMatrix(countData=countData,
                             colData=colData,
                             design=~condition)
dds = DESeq(dds)

```

Getting Results

## Data Visualization 

```{r}
res = results(dds, contrast=c("condition", "hoxa1_kd", "control_sirna"))

summary(res)
```

Making a Volcano Plot

```{r}
# Make a color vector for all genes
mycols <- rep("gray", nrow(res) )

# Color red the genes with absolute fold change above 2
mycols[ abs(res$log2FoldChange) > 2 ] <- "red"

# Color blue those with adjusted p-value less than 0.01
#  and absolute fold change more than 2
inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

plot( res$log2FoldChange, -log(res$padj), col=mycols, xlab="Log2(FoldChange)", ylab="-Log(P-value)" )
```

Adding Gene Annotation

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")

columns(org.Hs.eg.db)

res$symbol = mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column= "SYMBOL",
                    multiVals="first")


res$entrez = mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")

res$name =   mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype= "ENSEMBL",
                    column= "GENENAME",
                    multiVals="first")

head(res, 10)
```

```{r}
res = res[order(res$pvalue),]
write.csv(res, file="deseq_results.csv")
```

## Pathway Analysis 

```{r}
library(pathview)
library(gage)
library(gageData)

data(kegg.sets.hs)
data(sigmet.idx.hs)

kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs]
head(kegg.sets.hs, 3)
```

```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

Getting the results

```{r}
keggres = gage(foldchanges, gsets=kegg.sets.hs)
attributes(keggres)
head(keggres$less)
```

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110")
pathview(gene.data=foldchanges, pathway.id="hsa04110", kegg.native=FALSE)
## Focus on top 5 upregulated pathways here for demo purposes only
keggrespathways <- rownames(keggres$greater)[1:5]

# Extract the 8 character long IDs part of each string
keggresids = substr(keggrespathways, start=1, stop=8)
keggresids

```

```{r}
library(pathview)
library(knitr)

pic <- pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")


png_files <- list.files(pattern = "pathview\\.png$")
png_files 


include_graphics(png_files)
```

**\>Q**. Can you do the same procedure as above to plot the pathview figures for the top 5 down-reguled pathways?

## Gene Ontology tology 

```{r}
data(go.sets.hs)
data(go.subs.hs)

# Focus on Biological Process subset of GO
gobpsets = go.sets.hs[go.subs.hs$BP]

gobpres = gage(foldchanges, gsets=gobpsets, same.dir=TRUE)

lapply(gobpres, head)
```

#  Reactome Analysis

```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))
```

```{r}
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```

\>**Q**: What pathway has the most significant “Entities p-value”? Do the most significant pathways listed match your previous KEGG results? What factors could cause differences between the two methods?
