---
title: "Class07Lab"
author: "Peyton Chiu (PID:A18145937)"
format: gfm
editor: visual
---

# First up kmeans()

Demo od using kmeans() function in base R. First, make up some data with a known structure.

```{r}
tmp <- c( rnorm(30,-3), rnorm(30,3))
x <- cbind(tmp, rev(tmp))
plot(x)
```

Now we have some made up data in `x` let's see how kmeans works with this data

```{r}
k<- kmeans(x, centers =2, nstart =20)
k
```

> > Q. How many points are in each cluster

```{r}
k$size
```

> Q. How do we get to the cluster membership

```{r}
k$cluster
```

> Q. What about cluster centers

```{r}
k$centers 
```

Now we've got to the main results, let's use them to plot our data with the k-means result

```{r}
plot(x, col =k$cluster)
points(k$centers, col ="blue", pch=15)
```

## Now for hclust()

We will cluster the same data `x` with the `hclust()`. In this case `hclust()` requires a distance matrix input

```{r}

hc<- hclust(dist(x))
hc

```

Let's plot our hclust result

```{r}
plot(hc)
```

To get our cluster membership vector we need to "cut the tree with the `cutree()`

```{r}
grps<- cutree(hc, h=8)
grps
```

Now plot our data with the hclust() results

```{r}
plot(x, col =grps)
```

# Principal Component Analysis (PCA)

## PCA of UK food data

Read data from website and try a few visualizations

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names = 1)
x

```

> **Q1**. How many rows and columns are in your new data frame named `x`? What R functions could you use to answer this questions?

```{r}
dim(x)
```

There are the 17 rows and 4 columns

> **Q2.** Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

I prefer to use the `row.names=1` argument as it only has to add one line of code. Additionally, if you use the other approach and repeat you would end up corrupting the data as the first column would be remove, making you set the row names incorrectly.

```{r}
cols <- rainbow(nrow(x))
barplot(as.matrix(x), col=cols)
```

> **Q3**: Changing what optional argument in the above **barplot()** function results in the following plot?

Changing the besides argument in your code to FALSE results in the plot above. This because the different food types are stacked instead being shown side by side.

```{r}
barplot(as.matrix(x), col=cols, beside =TRUE)
```

```{r}
pairs(x, col=cols)
```

> **Q5**: Generating all pairwise plots may help somewhat. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

I can make some sense of the following code and resulting figure as I know that is creating scatterplots of food consumption for all possible countries against each other. If a given point lies on the diagonal for a given plot, that means the country's consumption of specific food item is similar within both countries.

> **Q6**. What is the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

Northern Ireland has more consumption of potatoes and less consumption of fresh fruit and Alcoholic drinks compared to the other countries based on this dataset.

PCA to the rescue

The main base R PCA function is called `prcomp()` and we will need to give it the transpose of our input data

```{r}
pca <- prcomp( t(x) )
summary(pca)
```

```{r}
attributes(pca)
```

To make our PCA plot (PCA score plot) we access `pca$X`

> **Q7**. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.

```{r}
pca$x
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(x))

```

color up the plot

> **Q8.** Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.

```{r}
country_cols <- c("orange", "red", "blue", "green")
plot(pca$x[,1],pca$x[,2], xlab = "PC1", ylab = "PC2")
text(pca$x[,1],pca$x[,2], colnames(x), col = country_cols)
```

```{r}
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v
```

```{r}
z <- summary(pca)
z$importance
```

```{r}
barplot(v, xlab="Principal Component", ylab="Percent Variation")
```

```{r}
par(mar = c(10, 3, 0.35, 0))
barplot(pca$rotation[,2], las = 2)


```

> **Q9**: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

The two food groups that feature prominently are Soft drinks and Fresh potatos. PC2 mainly tells us about differences between countries in their consumption of these foods a country with a high PC2 score will consume more Soft_drinks, while one with a low PC2 score will consume more Fresh_potatoes.

## PCA of RNA-Seq Data

Read data from website

```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
```

> **Q10**: How many genes and samples are in this data set?

```{r}
dim(rna.data)
```

There are 100 genes and 10 samples in this dataset

```{r}
pca<-prcomp(t(rna.data))
summary(pca)


```

Do our PCA plot of this RNA-seq data

```{r}
plot(pca$x[,1],pca$x[,2],xlab="PC1",ylab="PC2")
```

```{r}
plot(pca$x[,1],pca$x[,2],xlab="PC1",ylab="PC2")
text(pca$x[,1],pca$x[,2], colnames(rna.data))

```
