---
title: "Class 10 :Halloween Mini Project"
author: "Peyton Chiu (PID:A18145937)"
format: pdf
toc: true
---

## Background 

As it is nearly Halloween


## Data Import

```{r}
candy_file <- read.csv("candy-data.csv")
candy = data.frame(candy_file, row.names=1)
head(candy)
```

```{r}
flextable::flextable(head(candy,10))
```




>Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```
There are 85 candy types in this data set

>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```
There are 38 fruity candy types in the dataset 





```{r}
library(dplyr)

candy %>%
  filter(rownames(candy)=="Twix") %>%
  select(winpercent)
```


>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Swedish Fish",]$winpercent
```
My favorite candy in the data set is Swedish fish and its winpercent value is 54.86% 

>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```

The winpercent of Kit Kat is 76.76%

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

The winpercent for Tootsie Roll Snack Bars is 49.65%

## A quick overview of the dataset 

```{r}
library("skimr")
skim(candy)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, the variable for winpercent is on much higher scale than the other columns shown. 



>Q7. What do you think a zero and one represent for the candy$chocolate column?

That the candy does not contain chocolate 




>Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy, aes(x = winpercent)) +
  geom_histogram(bins = 30)
```


>Q9. Is the distribution of winpercent values symmetrical?

```{r}
library(ggplot2)

ggplot(candy, aes(x = winpercent)) +
  geom_density()
```

No the distribution is not symmetrical 

>Q10. Is the center of the distribution above or below 50

```{r}
summary(candy$winpercent)
```

The center of the distribution is lower than 50 as the median is 47.83


>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
#1 find all chocolate candy
#2 find their winpercent values
#3 calculate the mean of these values

#4-6 do the same for fruit candy 
#7 compare the means 
#8 pick the higher one

library(dplyr)
choco_candy <- candy %>% 
  filter(chocolate==1)

fruity_candy <- candy %>% 
  filter(fruity==1)

mean(choco_candy$winpercent)
mean(fruity_candy$winpercent)

```

On average chocolate candy is higher ranked than than fruit candy 

>Q12. Is this difference statistically significant

```{r}
t.test(choco_candy$winpercent,fruity_candy$winpercent)
```

These differences are statically significant 

>Q13. What are the five least liked candy types in this set?

```{r}
candy %>% arrange(winpercent) %>% head(5)
```

The five least liked candy types are Nik L lip, Boston Baked Beans, Chiclets, Super Bubble, Jawbusters 

>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>% arrange(desc(winpercent)) %>% head(5)
```

The top 5 all time favorite candy are Reese's Peanut Butter cup, Reese's miniatures, Twix, Kit Kat, Snickers 

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

library(ggplot2)

ggplot(candy, aes(winpercent,rownames(candy))) +
  geom_col() 

```


>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
library(ggplot2)

ggplot(candy, aes(winpercent, reorder(rownames(candy),winpercent))) +
  geom_col() 

```




> Q17. What is the worst ranked chocolate candy?

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols)
```

The worst ranked chocolate candy was Sixlets 


> Q18. What is the best ranked fruity candy

The best ranked fruit candy is star bursts


## Winpercent and Price Percent 

A plot with both variables/columns winpercent and pricepercent 

```{r}

library(ggrepel)

my_cols[as.logical(candy$fruity)] = "red"
ggplot(candy) +
  aes(winpercent,pricepercent, label =rownames(candy)) +
  geom_point(col=my_cols) +
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```


```{r}

```




>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

The candy type that is the highest ranked in terms of winpercent for the least money is Reeses miniatures 


>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

The 5 most expensive candies are Nik L Nip, Nestle Smarties,Ring pop	,Hershey's Krackel,Hershey's Milk Chocolate. Out of these the least popular one is Nik L Nip 



## Exploring the correlation structure 

Now that we’ve explored the dataset a little, we’ll see how the variables interact with one another. We’ll use correlation and view the results with the corrplot package to plot a correlation matrix.

```{r}
library(corrplot)

cij <-cor(candy)
corrplot(cij)
```


>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

The two variables that seem anti correlated are chocolate and fruity 



>Q23. Similarly, what two variables are most positively correlated?

The two variables that seem the most positively correlated are chocolate and winpercent 



## Principal Component Analysis 
The function to use this is called `prcomp()` with an optional scale argument . 

```{r}
pca <-prcomp(candy, scale =TRUE)
summary(pca)


```
Our main PCA results figure 
```{r}
ggplot(pca$x)+
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point(col=my_cols)+
  geom_text_repel(col=my_cols)
```

We should also examine the loadings or contributions of the original variables to the new PCs

```{r}
ggplot(pca$rotation)+
  aes(PC1,rownames(pca$rotation))+
  geom_col()
```


>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

The original variables that are picked up strongly by PC1 in the positive direction are fruity, hard, and pluribus. These makes sense as these characteristics are typically of fruity candies which usually comes in groups and are hard. 

Interactive plots that can be zoomed on and "brushed" over can be made with the **plotly** package. It's output is interactive and wil not render to pdf 
```{r}
p<-ggplot(pca$x)+
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point(col=my_cols)+
  geom_text_repel(col=my_cols)
```

```{r}
library(plotly)
##plotly(p)
```



